# Reproducible Reserach: Peer Assessment 1
## Oliver Crespo
## December 16, 2015

## Loading and processing the data

```{r activity chunk}
activity <- read.csv("activity.csv", colClasses = c("numeric", "character", "numeric"))
head(activity)
```

```{r}
names(activity)
```

```{r}
library(lattice)
activity$date <- as.Date(activity$date, "%Y-%m-%d")
```

## What is the mean number of steps taken per day?

First step is to aggregate the data
```{r aggregate}
Totalsteps <- aggregate(steps ~ date, data = activity, sum, na.rm = TRUE)
```

Now to create a nice graph:
```{r Total steps graph}
hist(Totalsteps$steps, main = "Total Steps by day", xlab = "day", col = "blue")
```

```{r meanandmedian}
mean(Totalsteps$steps)
median(Totalsteps$steps)
```


## What is the average daily activity pattern?

In order to make a time series of Type=1 and 5 min intervals:

```{r timeplot}
time_series <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
```

Now to plot it:

```{r plot}
plot(row.names(time_series), time_series, type = "l", xlab = "5-min interval", ylab = "Average across all Days", main = "Average number of steps taken", 
    col = "red")
```

The 5 minute interval with highsest number of stes on average is:
```{r}
max_interval <- which.max(time_series)
names(max_interval)
```


## Inputing missing values

In order to calculate and report the total number of missing values in the data (count how many NA's there are):
```{r NAs}
activity_NA <- sum(is.na(activity))
activity_NA
```

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval,etc

Fist Na replaced by mean in 5 min interval
```{r fill NAs}
StepsAverage <- aggregate(steps ~ interval, data = activity, FUN = mean)
fillNA <- numeric()
for (i in 1:nrow(activity)) {
    obs <- activity[i, ]
    if (is.na(obs$steps)) {
        steps <- subset(StepsAverage, interval == obs$interval)$steps
    } else {
        steps <- obs$steps
    }
    fillNA <- c(fillNA, steps)
}
```

Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r newdataset}
new_activity <- activity
new_activity$steps <- fillNA
```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r histogram}
Totalsteps2 <- aggregate(steps ~ date, data = new_activity, sum, na.rm = TRUE)
```
The Histogram is:
```{r}
hist(Totalsteps2$steps, main = "Total steps by day", xlab = "day", col = "blue")
```

and the mean is:
```{r}
mean(Totalsteps2$steps)
```

and a median of:
```{r}
median(Totalsteps2$steps)
```


## Are there differences in activity patterns between weekdyas and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
day <- weekdays(activity$date)
daylevel <- vector()
for (i in 1:nrow(activity)) {
    if (day[i] == "Saturday") {
        daylevel[i] <- "Weekend"
    } else if (day[i] == "Sunday") {
        daylevel[i] <- "Weekend"
    } else {
        daylevel[i] <- "Weekday"
    }
}
activity$daylevel <- daylevel
activity$daylevel <- factor(activity$daylevel)

stepsByDay <- aggregate(steps ~ interval + daylevel, data = activity, mean)
names(stepsByDay) <- c("interval", "daylevel", "steps")
```
Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data:

```{r}
xyplot(steps ~ interval | daylevel, stepsByDay, type = "l", layout = c(1, 2), 
    xlab = "Interval", ylab = "Number of steps")
```
